GSE158420: RNA expression downsampling

### 2024-07-17 analysis/GSE158420_rna.Rmd
fix figure
plotting

fres <- lapply(sizes, function(size) {
  sigup <- head(rownames(subset(dge,padj<0.05 & log2FoldChange>0)),size)
  sigdn <- head(rownames(subset(dge,padj<0.05 & log2FoldChange<0)),size)
  bg <- rownames(dge)

@@

fres <- do.call(rbind,fres)
rownames(fres) <- sizes
colnames(fres) <- c("up","dn")
fres <- as.data.frame(fres)
fres$tot <- fres$up + fres$dn
tot <- fres$tot
names(tot) <- sizes
par(mar=c(mar = c(5.1, 5.1, 4.1, 2.1)))
barplot(tot,horiz=TRUE,las=1,xlab="no. pathways FDR<0.05",ylab="no. genes selected")

@@

SAMPLESIZES=c(2,3,6,10,16,22,30)
#SAMPLESIZES=c(3,6,10,22)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100) # 50 repetitions
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    ss2 <- ss[mysample,]
    x2 <- x[,mysample]
    x2 <- x2[rowMeans(x2)>=10,]

@@

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

pdf("sens_fgsea.pdf",height=7,width=5)

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

@@

SAMPLESIZES=c(2,3,6,10,16,22,30)
ssres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100) # 50 repetitions
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    ss2 <- ss[mysample,]
    x2 <- x[,mysample]
    x2 <- x2[rowMeans(x2)>=10,]

@@

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,nrow(fsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(fsig))

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

pdf("sens_fora.pdf",height=7,width=5)

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,nrow(fsig))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=nrow(fsig))

@@

## Session information

>>>>

ores <- lapply(sizes, function(size) {
  sigup <- head(rownames(subset(dge,padj<0.05 & log2FoldChange>0)),size)
  sigdn <- head(rownames(subset(dge,padj<0.05 & log2FoldChange<0)),size)
  bg <- rownames(dge)

@@

ores <- do.call(rbind,ores)
rownames(ores) <- sizes
colnames(ores) <- c("up","dn")
ores <- as.data.frame(ores)
ores$tot <- ores$up + ores$dn
tot <- fres$tot
names(tot) <- sizes
par(mar=c(mar = c(5.1, 5.1, 4.1, 2.1)))
barplot(tot,horiz=TRUE,las=1,xlab="no. pathways FDR<0.05",ylab="no. genes selected")

@@

SAMPLESIZES=c(2,3,5,10,15,20,25,30)
#SAMPLESIZES=c(3,6,10,22)
fgres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100) # 50 repetitions
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    ss2 <- ss[mysample,]
    x2 <- x[,mysample]
    x2 <- x2[rowMeans(x2)>=10,]

@@

ssres <- fgres
par(mfrow=c(3,1))

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

pdf("sens_fgsea.pdf",height=7,width=5)
par(mfrow=c(3,1))

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

ssres <- fgres
dots <- lapply(ssres,function(x) { x[,"TP"] })
fgres2 <- do.call(rbind,lapply(dots,summary))

@@

SAMPLESIZES=c(2,3,5,10,15,20,25,30)
fores <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100) # 50 repetitions
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    ss2 <- ss[mysample,]
    x2 <- x[,mysample]
    x2 <- x2[rowMeans(x2)>=10,]

@@

ssres <- fores

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,length(osets))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=length(osets))

@@

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
lapply(ratio,summary)
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion inconsistent",xlab="sample size",cex=0,ylim=c(0,0.5)) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

pdf("sens_fora.pdf",height=7,width=5)
par(mfrow=c(3,1))

dots <- lapply(ssres,function(x) { x[,"TP"] })
lapply(dots,summary)
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. consistent",xlab="sample size",cex=0,main="fora",ylim=c(0,length(osets))) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h=length(osets))

@@

## Summarise downnsampling

```{r,downsample,fig.width=10}

par(mfrow=c(1,3))

dots <- lapply(fgres,function(x) { x[,"TP"] })
fgres2 <- do.call(rbind,lapply(dots,summary))

dots <- lapply(fores,function(x) { x[,"TP"] })
fores2 <- do.call(rbind,lapply(dots,summary))

df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
colnames(df) <- c("Sample size", "fora", "fgsea")

MAX=length(fsets)
plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", ylim=c(0,MAX))
points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
abline(h=length(osets),col="red")
abline(h=length(fsets),col="blue")
mtext("no. consistent pathways FDR<0.05")
grid()

legend("bottomright", legend=c("fgsea","fora"),
       col=c("blue","red"), lty=1, pch=19,  cex=1)

dots <- lapply(fgres,function(x) { x[,"FP"] })
fgres2 <- do.call(rbind,lapply(dots,summary))

dots <- lapply(fores,function(x) { x[,"FP"] })
fores2 <- do.call(rbind,lapply(dots,summary))

df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
colnames(df) <- c("Sample size", "fora", "fgsea")

MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",ylim=c(0,MAX))
points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
mtext("no. inconsistent pathways FDR<0.05")
grid()

dots <- lapply(fores,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(fores,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
fores2 <- do.call(rbind,lapply(ratio,summary))

dots <- lapply(fgres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(fgres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
fgres2 <- do.call(rbind,lapply(ratio,summary))

df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
colnames(df) <- c("Sample size", "fora", "fgsea")

MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",ylim=c(0,MAX))
points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
mtext("Proportion of inconsistent pathways")
grid()

### PDF
pdf("sens.pdf",height=4,width=8)
par(mfrow=c(1,3))

dots <- lapply(fgres,function(x) { x[,"TP"] })
fgres2 <- do.call(rbind,lapply(dots,summary))

dots <- lapply(fores,function(x) { x[,"TP"] })
fores2 <- do.call(rbind,lapply(dots,summary))

df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
colnames(df) <- c("Sample size", "fora", "fgsea")

MAX=length(fsets)
plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways", main="no. consistent pathways FDR<0.05",ylim=c(0,MAX))
points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
abline(h=length(osets),col="red")
abline(h=length(fsets),col="blue")
grid()

legend("bottomright", legend=c("fgsea","fora"),
       col=c("blue","red"), lty=1, pch=19,  cex=1)

dots <- lapply(fgres,function(x) { x[,"FP"] })
fgres2 <- do.call(rbind,lapply(dots,summary))

dots <- lapply(fores,function(x) { x[,"FP"] })
fores2 <- do.call(rbind,lapply(dots,summary))

df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
colnames(df) <- c("Sample size", "fora", "fgsea")

MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="no. pathways",main="no. inconsistent pathways FDR<0.05",ylim=c(0,MAX))
points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
grid()

dots <- lapply(fores,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(fores,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
fores2 <- do.call(rbind,lapply(ratio,summary))

dots <- lapply(fgres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(fgres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
fgres2 <- do.call(rbind,lapply(ratio,summary))

df <- data.frame(SAMPLESIZES,fores2[,"Median"],fgres2[,"Median"])
colnames(df) <- c("Sample size", "fora", "fgsea")

MAX=max(c(fores2[,"Median"],fgres2[,"Median"]))
plot(df$`Sample size`,df$fora,type="b",col="red",pch=19,xlab="Sample size",ylab="proportion",main="Proportion of inconsistent pathways",ylim=c(0,MAX))
points(df$`Sample size`,df$fgsea,type="b",col="blue",pch=19 )
grid()
dev.off()

```
## Session information
